---
output:
  rmarkdown::github_document:
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fashr

<!-- badges: start -->
<!-- badges: end -->

The goal of fashr is to implement the functional adaptive shrinkage through empirical Bayes.

**Note:** The code to generate the results for the manuscript is [here](https://github.com/stephenslab/fashr-paper).

## Installation

You can install the development version of fashr from [GitHub](https://github.com/) with:

```{r}
# install.packages("pak")
pak::pak("stephenslab/fashr")
```

## Example

This is a basic example, which tries to identify non-linear functions in a set of datasets.
We use `order = 2` to apply the second order IWP, which has a base model S<sub>0</sub> being the space of linear functions.

```{r example, results="hide"}
library(fashr)

# Simulate five datasets
set.seed(123)
datasets <- list()
for (i in 1:20) {
  n <- 50
  t <- seq(0, 5, length.out = n)
  sd <- sample(c(2, 1), size = n, replace = TRUE)
  u <- runif(1); if (u < 0.5) {f <- function(t) 3*cos(t)} else {f <- function(t) (t)}
  y <- f(t) + rnorm(n, sd = 0.5)
  datasets[[i]] <- data.frame(t = t, y = y, sd = sd)
}

# Fit the model
fash_fit <- fash(Y = "y", smooth_var = "t", S = "sd", data = datasets, 
                  order = 2, likelihood = "gaussian", verbose = TRUE)
```

```{r}
fash_fit
```

Take a look at the structure plot ordered by local false discovery rate:

```{r, fig.height=3.5, fig.width=4}
plot(fash_fit, ordering = "lfdr")
```

Obtain the posterior summary of the function for the first dataset:

```{r}
fitted <- predict(fash_fit, index = 1)
str(fitted)
```

Obtain the posterior samples of the function:

```{r}
fitted_samps <- predict(fash_fit, index = 1, only.samples = TRUE, M = 30)
str(fitted_samps)
```

```{r, fig.height=4, fig.width=4}
plot(datasets[[1]]$t, datasets[[1]]$y, type = "p", col = "black", ylab = "y", xlab = "t")
lines(fitted$x, fitted$mean, col = "red")
polygon(c(fitted$x, rev(fitted$x)), c(fitted$lower, rev(fitted$upper)), col = rgb(1, 0, 0, 0.2), border = NA)
matlines(fitted$x, fitted_samps[, 1:5], col = "blue", lty = 2, lwd = 0.5)
```

Compute the FDR, and highlight the datasets with FDR < 0.1:

```{r, fig.height=5, fig.width=5}
fdr_result <- fdr_control(fash_fit, alpha = 0.1, plot = TRUE)
```

```{r}
str(fdr_result)
```
